Iterative Reduction: Parsing and Interpreting Code Without Trees

### 2025-01-21 Paper/paper.tex
undo
new comments

We propose a novel method for the process of interpretation, which doesn't use parse trees.
But our algorithm is not limited to this, and can be altered to produce parse trees as well (see $\S$\ref{producing-parse-trees}).
This means that our algorithm can be utilized as part of the pipeline for compilers and traditional interpreters.

@@

We can formalize the algorithm given in the introduction as follows: we start with an alphabet of tokens $\Sigma$.
Define $\oS=\Sigma\times\bbN$, and we assume the existence of a so-called \textit{initial $\beta$-reducer}:
$\hb\colon \Sigma\times\Sigma \longto \Sigma$, which is defined by the implementer of the language being interpreted, and is the function which
tells us how to combine tokens together.
We extend it to a \textit{derived $\beta$-reducer} which is a function $\beta\colon\oS^*\longto\oS^*$ as follows.

Suppose $\xi=\sigma^1_{p_1}\cdots\sigma^n_{p_n}$, then $\beta(\xi)$ is defined recursively as follows:
\begin{enumerate}
    \item if $n=1$, then $\beta(\xi)=\sigma^n_0$,
    \item if $p_1\geq p_2$, then $\beta(\xi)=\hb(\sigma^1,\sigma^2)_{p_2}\sigma^3_{p_3}\cdots\sigma^n_{p_n}$,
    \item otherwise $\beta(\xi)=\sigma^1_{p_1}\,\beta(\sigma^2_{p_2}\cdots\sigma^n_{p_n})$.
\end{enumerate}

@@

    We define our alphabet $\Sigma$ to be the same as in~\ref{sec:firstattempt}.
    Now we define our initial $\beta$-reducer which will support parentheses.
    It is an extension of our previous $\beta$-reducer:
    \begin{gather*}
        \hb(\mbt x,\mb\circ) = (\mb{\ttt x,\circ},\snd),\quad \hb(\mb{\ttt x,\circ},\mbt y) = (\mb{\ttt x\circ\ttt y},\snd) \qquad \hbox{for $\circ\in\set{\ttt{+,*}}$} \\
        \hb(\mbt{x,*},\mbt{y,+}) = (\mbt{x*y,+},\snd)
    \end{gather*}
    where $\fst,\snd$ are the first and second projections, respectively (i.e. $(x,y)\mapsto x$ and $(x,y)\mapsto y$).
    We then add the rules for parentheses:
    \begin{gather*}
        \hb(\mbt x,\mbt)) = (\mbt{x,)},\snd),\quad \hb{\mb{\ttt x,\circ},\mbt{y,)}} = (\mb{\ttt x\circ\ttt y,\ttt)},\snd) \qquad \hbox{for $\circ\in\set{\ttt{+,*}}$} \\
        \hb(\mbt(,\mbt{x,)}) = (\mbt x,\fst)
    \end{gather*}
    \gobble) % syntax highlighting
    And we define our priority function as follows:
    \yoni{feels out of context, the notion of a priority function
    was not defined. Also, notice that some tokens get priorities and some don't, which is a bit confusing.}
    \ari{Both points are now explained in the previous section, is this good?}
    \[ \pi(\mbt x) = \infty\hbox{ for numbers \ttt x},\quad \pi(\mbt+) = 1,\quad \pi(\mbt*) = 2,\quad \pi(\mbt() = \infty,\quad \pi(\mbt)) = 0 \]

@@

@@

Here we abstract away the value of each term; the value of the new term given by the initial $\beta$-reducer will be determined by an output function.
Our alphabet of tokens $\Sigma$, instead of having a token for every number, will just have tokens for types (number, operator, etc.).
We also have another set $\values$, the \textit{universe of values}.

@@

The initial $\beta$-reducer is a function
\[ \hb \colon \Sigma\times\Sigma \partial \Sigma\times(\bbZ\times\bbZ\to\bbZ)\times(\values\times\values\partial\values) \]
Define $\oP=\oS\times\values$ to be the set of all possible terms,
that is, 
a term consists of a type, a priority, and a value.
Since terms now consist of three parts, we write them as $\sigma_p(v)$, where $\sigma$ is the type; $p$ is the priority; and $v$ the value.

@@

    Define $\Sigma=\set{\NUM,\OP,\LPAREN,\RPAREN,\OP\NUM,\RPAREN\NUM}$.
    Our initial $\beta$-reducer will is defined as follows:
    \begin{itemize}
        \item $\hb(\NUM,\OP)=(\OP\NUM,\snd,(v,f\mapsto(v,f))$
        \item $\hb(\OP\NUM,\NUM)=(\NUM,\snd,((v,f),u\mapsto f(v,u)))$
        \item $\hb(\OP\NUM,\OP\NUM)=(\OP\NUM,\snd,((v,f),(u,g)\mapsto (f(v,u),g))$
        \item $\hb(\NUM,\RPAREN)=(\RPAREN\NUM,\snd,(v,\_\mapsto v))$
        \item $\hb(\OP\NUM,\RPAREN\NUM)=(\RPAREN\NUM,\snd,((v,f),u\mapsto f(v,u)))$
        \item $\hb(\LPAREN,\RPAREN\NUM)=(\NUM,\fst,(\_,v\mapsto v))$
    \end{itemize}
    In the first point $(v,f\mapsto(v,f))$ means that the function to compute the value of the new token is one which takes two values $v,f$ and outputs a new value which is the pair $(v,f)$.

    So taking our previous example of \ttt{2*(1+3)+4}: we first must convert it into tokens acceptable for $\beta$-reduction (via lexing).
    In this example allocating initial priorities is not so simple: the priorities depend not on the type of the token in $\Sigma$, but the token in the stream given to the lexer.
    So we will assume the lexer allocates priorities similar to the previous example in the previous section.
    This issue will be resolved in the next section.
    So the stream is converted to
    \[ \NUM_\infty(2)\ \OP_2(*)\ \LPAREN_\infty\ \NUM_\infty(1)\ \OP_1(+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \]
    We no longer need boxes to distinguish between tokens because the individual tokens and so we will no longer use them.

    The derivation is then
    \begin{gather*}
        \OP\NUM_2(2,*)\ \LPAREN_\infty\ \NUM_\infty(1)\ \OP_1(+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \\
        \OP\NUM_2(2,*)\ \LPAREN_\infty\ \OP\NUM_1(1,+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \\
        \OP\NUM_2(2,*)\ \LPAREN_\infty\ \OP\NUM_1(1,+)\ \RPAREN\NUM_0(3)\ \OP_1(+)\ \NUM_\infty(4) \\
        \OP\NUM_2(2,*)\ \LPAREN_\infty\ \RPAREN\NUM_0(4)\ \OP_1(+)\ \NUM_\infty(4) \\
        \OP\NUM_2(2,*)\ \NUM_\infty(4)\ \OP_1(+)\ \NUM_\infty(4) \\
        \OP\NUM_2(2,*)\ \OP\NUM_1(4,+)\ \NUM_\infty(4) \\
        \OP\NUM_2(8,+)\ \NUM_\infty(4) \\
        \OP\NUM_2(8,+)\ \NUM_0(4) \\
        \NUM_0(12)
    \end{gather*}

@@

Finally, our \textit{initial $\beta$-reducer} is a function ($\state$ is a set of functions $\termp\partial\expi$ which we define later)
\yoni{Using something and defining it later should be avoided...}
\ari{We can move the definition up, but I felt like it messed with the flow.}

\yoni{Generally, this section is hard to follow. I suggest adding plenty of examples, after every new definition / concept, please add an appropriate small example.}
\ari{I'm not really sure what kinds of examples to add; I added a big example at the end that we can potentially split up?
But I feel like we need all of the definitions to give a good example.}

>>>>

We propose a novel method for the process of interpretation, which doesn't use parse trees.
But our algorithm is not limited to this, and can be altered to produce parse trees as well (see

%%\S\ref{producing-parse-trees}).
This means that our algorithm can be utilized as part of the pipeline for compilers and traditional interpreters.

@@

We can formalize the algorithm given in the introduction as follows: we start with an alphabet of tokens $\Sigma$.
Define $\oS=\Sigma\times\bbN$, and we assume the existence of a so-called \textit{initial $\beta$-reducer}:
$\hb\colon \Sigma\times\Sigma \longto \Sigma$, which is defined by the implementer of the language being interpreted, and is the function which
tells us how to combine tokens together.
We extend it to a \textit{derived $\beta$-reducer} which is a function $\beta\colon\oS^*\longto\oS^*$ as follows.
\yoni{plz explicitly say that we use $a_b$ to denote $<a,b>$.}
Suppose $\xi=\sigma^1_{p_1}\cdots\sigma^n_{p_n}$, then $\beta(\xi)$ is defined recursively as follows:
\begin{enumerate}
    \item if $n=1$, then $\beta(\xi)=\sigma^n_0$,
    \item if $p_1\geq p_2$, then $\beta(\xi)=\hb(\sigma^1,\sigma^2)_{p_2}\sigma^3_{p_3}\cdots\sigma^n_{p_n}$,
    \item otherwise $\beta(\xi)=\sigma^1_{p_1}\,\beta(\sigma^2_{p_2}\cdots\sigma^n_{p_n})$.
\end{enumerate}

\yoni{I wonder how can this support ternary operators like if then else, integrals, or just custom-defined functions. Maybe like in functional programming, so that functions will also be values?}

@@

    We define our alphabet $\Sigma$ to be the same as in~\ref{sec:firstattempt}.
    Now we define our initial $\beta$-reducer which will support parentheses.
    It is an extension of our previous $\beta$-reducer:
    \begin{gather*}
        \hb(\mbt x,\mb\circ) = (\mb{\ttt x,\circ},\snd),\quad \hb(\mb{\ttt x,\circ},\mbt y) = (\mb{\ttt x\circ\ttt y},\snd) \qquad \hbox{for $\circ\in\set{\ttt{+,*}}$} \\
        \hb(\mbt{x,*},\mbt{y,+}) = (\mbt{x*y,+},\snd)
    \end{gather*}
    where $\fst,\snd$ are the first and second projections, respectively (i.e. $(x,y)\mapsto x$ and $(x,y)\mapsto y$).
    We then add the rules for parentheses:
    \begin{gather*}
        \hb(\mbt x,\mbt)) = (\mbt{x,)},\snd),\quad \hb{\mb{\ttt x,\circ},\mbt{y,)}} = (\mb{\ttt x\circ\ttt y,\ttt)},\snd) \qquad \hbox{for $\circ\in\set{\ttt{+,*}}$} \\
        \hb(\mbt(,\mbt{x,)}) = (\mbt x,\fst)
    \end{gather*}
    \gobble) % syntax highlighting
    And we define our priority function as follows:
    \[ \pi(\mbt x) = \infty\hbox{ for numbers \ttt x},\quad \pi(\mbt+) = 1,\quad \pi(\mbt*) = 2,\quad \pi(\mbt() = \infty,\quad \pi(\mbt)) = 0 \]

@@

\yoni{It would be nice to really show what would have happened if we were to use the previous section's approach. The main difference, if i understand correctly, is that this would enforce $\hb(\mb (, \mb{x,)}$ to be $(x,snd)$ instead of $(x,fst)$, right? It would be cool to write out which term we would eventually get this way.}

@@

Here we abstract away the value of each term; the value of the new term given by the initial $\beta$-reducer will be determined by an output function.
Our alphabet of tokens $\Sigma$, instead of having a token for every number, will just have tokens for types (number, operator, etc.).
We also have another set $\values$, the \textit{universe of values}.
\yoni{plz provide an example for what this universe typically looks like.}

@@

The initial $\beta$-reducer is a function
\[ \hb \colon \Sigma\times\Sigma \partial \Sigma\times(\bbZ\times\bbZ\to\bbZ)\times(\values\times\values\partial\values) \]

\yoni{plz intuitively describe what it returns.}

Define $\oP=\oS\times\values$ to be the set of all possible terms,
that is, 
a term consists of a type, a priority, and a value.
\yoni{A term is a common word for something like an expression, separated from priority and type (which is here not really a type). So this naming is problematic.}
Since terms now consist of three parts, we write them as $\sigma_p(v)$, where $\sigma$ is the type; $p$ is the priority; and $v$ the value.
\yoni{Not sure about value either. Let's discuss. Value is typically a number, a string, a concrete function...}

@@

    Define $\Sigma=\set{\NUM,\OP,\LPAREN,\RPAREN,\OP\NUM,\RPAREN\NUM}$.
    Our initial $\beta$-reducer is defined as follows:
    \begin{itemize}
        \item $\hb(\NUM,\OP)=(\OP\NUM,\snd,(v,f\mapsto(v,f))$
\yoni{This is a bit strange. It suggests that $\values\times \values \subseteq \values$, which is impossible. We need to properly define the relevant set here, and the domain and range of this function.}
        \item $\hb(\OP\NUM,\NUM)=(\NUM,\snd,((v,f),u\mapsto f(v,u)))$
        \item $\hb(\OP\NUM,\OP\NUM)=(\OP\NUM,\snd,((v,f),(u,g)\mapsto (f(v,u),g))$
        \item $\hb(\NUM,\RPAREN)=(\RPAREN\NUM,\snd,(v,\_\mapsto v))$
        \item $\hb(\OP\NUM,\RPAREN\NUM)=(\RPAREN\NUM,\snd,((v,f),u\mapsto f(v,u)))$
        \item $\hb(\LPAREN,\RPAREN\NUM)=(\NUM,\fst,(\_,v\mapsto v))$
    \end{itemize}
\yoni{Here would be a good place to show what is saved compared to the previous approach, maybe there are less lines or something?}
    In the first point $(v,f\mapsto(v,f))$ means that the function to compute the value of the new token is one which takes two values $v,f$ and outputs a new value which is the pair $(v,f)$.

    So taking our previous example of \ttt{2*(1+3)+4}: we first must convert it into tokens acceptable for $\beta$-reduction (via lexing).
    In this example allocating initial priorities is not so simple: the priorities depend not on the type of the token in $\Sigma$, but the token in the stream given to the lexer.
    So we will assume the lexer allocates priorities similar to the previous example in the previous section.
    \yoni{better to use ref instead of "previous example".}
    This issue will be resolved in the next section.
    So the stream is converted to
    \[ \NUM_\infty(2)\ \OP_2(*)\ \LPAREN_\infty\ \NUM_\infty(1)\ \OP_1(+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \]
    We no longer need boxes to distinguish between tokens because the individual tokens and so we will no longer use them.
    \yoni{"and so" sounds like it misses something before it.}

    The derivation is then
    \begin{gather*}
        \OP\NUM_2(2,*)\ \LPAREN_\infty\ \NUM_\infty(1)\ \OP_1(+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \\
        \OP\NUM_2(2,*)\ \LPAREN_\infty\ \OP\NUM_1(1,+)\ \NUM_\infty(3)\ \RPAREN_0\ \OP_1(+)\ \NUM_\infty(4) \\
        \OP\NUM_2(2,*)\ \LPAREN_\infty\ \OP\NUM_1(1,+)\ \RPAREN\NUM_0(3)\ \OP_1(+)\ \NUM_\infty(4) \\
        \OP\NUM_2(2,*)\ \LPAREN_\infty\ \RPAREN\NUM_0(4)\ \OP_1(+)\ \NUM_\infty(4) \\
        \OP\NUM_2(2,*)\ \NUM_\infty(4)\ \OP_1(+)\ \NUM_\infty(4) \\
        \OP\NUM_2(2,*)\ \OP\NUM_1(4,+)\ \NUM_\infty(4) \\
        \OP\NUM_2(8,+)\ \NUM_\infty(4) \\
        \OP\NUM_2(8,+)\ \NUM_0(4) \\
        \NUM_0(12)
    \end{gather*}
\yoni{$\OP\NUM_2(2,*)$ should be $\OP\NUM_2((2,*))$ right?}

@@

Finally, our \textit{initial $\beta$-reducer} is a function ($\state$ is a set of functions $\termp\partial\expi$ which we define later)
\yoni{Using something and defining it later should be avoided...}
\ari{We can move the definition up, but I felt like it messed with the flow.}
\yoni{But also using something and defining it later messes the flow... We can probably find a way that is clear without jumps.}

\yoni{Generally, this section is hard to follow. I suggest adding plenty of examples, after every new definition / concept, please add an appropriate small example.}
\ari{I'm not really sure what kinds of examples to add; I added a big example at the end that we can potentially split up?
But I feel like we need all of the definitions to give a good example.}
\yoni{exactly -- the idea is to break the example to many small examples, so that each concept is clearer.}
